RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-05-24 docs/paper/paper.tex
docs: Tweak minor issues

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress. Initially, the piece consists of a user-defined cantus firmus, the first note of counterpoint line, and the last note of counterpoint line. An action is extension of the counterpoint left part by one new note. Not every action is valid, there are rules prohibiting some of them. Invalid actions can not be done by design. When the last note is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.

@@

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search (VNS) is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. Despite that, there are important differences:
\begin{itemize}
	\item In \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, there are only evaluation rules and there are no rules of composition. If a piece breaks some fundamental rules of counterpoint, it is still evaluated (but it can receive score making it unfeasible). Conversely, in the current study, both rules of composition and rules of evaluation are present as separate entities. An action that contradicts a rule of composition can not be done and so only compliant pieces are evaluated.
	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally. From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sounds unpleasant, because some of these prescriptions are style-specific.
	\item Output of VNS is a fixed piece. If cantus firmus is changed, VNS should be run from scratch. Actually, output of Monte Carlo Beam Search is a fixed piece too and search should be run from scratch for every new cantus firmus as well. However, the current approach can be upgraded to a true reinforcement learning approach. Such upgrade is possible, because the problem is formulated in RL terms: composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions updates. If so, there can be an agent that observes current state of a piece and writes counterpoint line based on observations. Cantus firmus can be changed from episode to episode during training of the agent in order to teach the agent to deal with various cantus firmi. After each episode played by the trained agent, a new output piece can be created even for a fixed cantus firmus.
	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results can not be reproduced. Unlike this, links from the present paper lead to time-tested sites such as GitHub and PyPI.
\end{itemize}

@@

To start with, define representation of a musical piece. Suppose that a diatonic scale (like C-major or A-minor) is chosen. Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eighths of measure). In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint. Definition of cantus firmus implies that every its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus. Denote this length by $m$, where $m$ is usually between 8 and 24. From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition. Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.

@@

Here, only one rule of rhythm is present:
\begin{itemize}
	\item There must be no conflicts between strong beats and longer notes. In other words, every measure must be split in a manner that allows a listener to unambiguously define meter. For example, if a measure is split like this: (4, 2, 1, 1) or (2, 1, 1, 8) where 8 stands for a tied over bar note, this is correct, but if a measure is split like this: (2, 4, 2), this is not correct, because downbeat note is shorter than the next note and intermediate strong beat is inside a span of a note, not at a beginning of a note.
\end{itemize}
Further, seven rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item No pitches can be rearticulated two times in a row, because it makes a line stalling;
	\item A line can not move from an intermediate pitch by more than 9 semitones without any changes in direction (rearticulation is not a change of direction);
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item There can be no more than 2 skips (leaps) in a row;
	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
	\item For every note from counterpoint line, there must be a way to reach the final note of this line with step motion (remember that the final note is set initially), because resolution must be present during last measures.
\end{itemize}
Finally, there are seven rules of harmony:
\begin{itemize}
	\item If a note from a counterpoint line starts on a strong beat (downbeat or measure middle), harmonic interval between the pitch and simultaneously sounding pitch from cantus firmus, must be consonant (although, if counterpoint note is suspended, it can be dissonant with the second simultaneously sounding note from cantus firmus);
	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be reached by step motion;
	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be left by step motion;
	\item If a dissonance is created by suspended note, this dissonance must be resolved by downward step motion;
	\item Harmonic intervals larger than tenth are prohibited;
	\item Voice crossing is prohibited (i.e., counterpoint can be either strictly above cantus firmus or strictly below it);
	\item Overlapping motion is prohibited (i.e., counterpoint can be either strictly above lagged cantus firmus or strictly below it).
\end{itemize}

>>>>

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress. Initially, the piece consists of a user-defined cantus firmus, the first note of counterpoint line, and the last note of counterpoint line. An action is extension of the counterpoint left part by one new note. Not every action is valid, there are rules prohibiting some of them. Invalid actions cannot be done by design. When the last note is reached, the episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.

@@

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search (VNS) is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. Despite that, there are important differences:
\begin{itemize}
	\item In \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, there are only evaluation rules and there are no rules of composition. If a piece breaks some fundamental rules of counterpoint, it is still evaluated (but it can receive score making it unfeasible). Conversely, in the current study, both rules of composition and rules of evaluation are present as separate entities. An action that contradicts a rule of composition cannot be done and so only compliant pieces are evaluated.
	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally. From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sound unpleasant, because some of these prescriptions are style-specific.
	\item Output of VNS is a fixed piece. If cantus firmus is changed, VNS should be run from scratch. Actually, output of Monte Carlo Beam Search is a fixed piece too and search should be run from scratch for every new cantus firmus as well. However, the current approach can be upgraded to a true reinforcement learning approach. Such upgrade is possible, because the problem is formulated in RL terms: composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions updates. If so, there can be an agent that observes current state of a piece and writes counterpoint line based on observations. Cantus firmus can be changed from episode to episode during training of the agent in order to teach the agent to deal with various cantus firmi. After each episode played by the trained agent, a new output piece can be created even for a fixed cantus firmus.
	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results cannot be reproduced. Unlike this, links from the present paper lead to time-tested sites such as GitHub and PyPI.
\end{itemize}

@@

To start with, define representation of a musical piece. Suppose that a diatonic scale (like C-major or A-minor) is chosen. Let a line be a list of $(\mathrm{pitch}, \mathrm{duration})$ pairs where $\mathrm{pitch}$ must be from the scale (i.e., only 7 out of 12 notes per octave can be used and so chromaticism is prohibited) and $\mathrm{duration} \in \{1, 2, 4, 8\}$ (i.e., it is measured in eighths of measure). In this study, pieces are written exactly in 2 melodic lines (parts, voices) where one of them is cantus firmus and the other one is counterpoint. Definition of cantus firmus implies that each its element has duration equal to 8 and duration of a piece in measures is equal to length of cantus firmus. Denote this length by $m$, where $m$ is usually between 8 and 24. From music point of view, this means that a piece to be created is rather a phrase that can be inserted into a longer composition. Such limitation goes back to species counterpoint, because it is designed exactly for writing short phrases.

@@

Here, only one rule of rhythm is present:
\begin{itemize}
	\item There must be no conflicts between strong beats and longer notes. In other words, every measure must be split in a manner that allows a listener to unambiguously define meter. For example, if a measure is split like this: (4, 2, 1, 1) or (2, 1, 1, 8) where 8 stands for a tied over bar note, this is correct, but if a measure is split like this: (2, 4, 2), this is not correct, because downbeat note is shorter than the next note and intermediate strong beat is inside a span of a note, not at a beginning of a note.
\end{itemize}
Further, seven rules of voice leading are imposed:
\begin{itemize}
	\item Only pitches from tonic triad (i.e., tonic, mediant, and dominant) can be rearticulated (repeated), because only they are stable;
	\item No pitches can be rearticulated two times in a row, because it makes a line stalling;
	\item A line cannot move from an intermediate pitch by more than 9 semitones without any changes in direction (rearticulation is not a change of direction);
	\item If a line has submediant followed by leading tone, tonic must be used after leading tone, because there is strong attraction to it; similarly, if a line has leading tone followed by submediant, dominant must be used after submediant;
	\item There can be no more than 2 skips (leaps) in a row;
	\item Movement by step in the opposite direction must happen immediately after a skip larger than third in order to resolve tension associated with the skip;
	\item For every note from counterpoint line, there must be a way to reach the final note of this line with step motion (remember that the final note is set initially), because resolution must be present during last measures.
\end{itemize}
Finally, there are seven rules of harmony:
\begin{itemize}
	\item If a note from a counterpoint line starts on a strong beat (downbeat or measure middle), harmonic interval between the pitch and simultaneously sounding pitch from cantus firmus, must be consonant (although, if counterpoint note is suspended, it can be dissonant with the second simultaneously sounding note from cantus firmus);
	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be reached by step motion;
	\item If a note from a counterpoint line forms a dissonance with cantus firmus, it must be left by step motion;
	\item If a dissonance is created by suspended note, this dissonance must be resolved by downward step motion;
	\item Harmonic intervals larger than tenth are prohibited;
	\item Voice crossing is prohibited (i.e., counterpoint can be either strictly above cantus firmus or strictly below it);
	\item Overlapping motion is prohibited (i.e., counterpoint can be either strictly above lagged cantus firmus or strictly below it).
\end{itemize}
